Epistemic Semantics in Guarded String Models

### 2020-09-25 tex/sdraft.tex
add modal axioms

\newcommand{\ehc}[1]{\textit{\textcolor{green}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}

@@

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o1  h:h \\
            event o0  t:t \\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent aly \\
            \hspace{3mm} o1  -> o1 \\
            \hspace{3mm} o0  -> o0 \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

@@

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{ll}
      \text{events} & e \in \mathsf{P} \\
      \text{states} & \sigma~\text{as in Figure 1} \\    
      p,q & ::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap \\
      \end{array}\)

      \vspace{1mm}
      \begin{tabular}{cl}
        $\Box_a p$ & $\triangleq \neg\Diamond_a\neg p$ \\
        \(p \wedge q\) &  \(\triangleq \neg(\neg p + \neg q)\) \\
        $\bullet$ & \( \triangleq \sum_{e \in \mathsf P}e\)
      \end{tabular}
  \end{minipage}}

@@

\begin{example}   \label{interpretation1}
\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
$\Box_a \bullet^* h$  & Amy knows that it's heads. \\
\end{tabular}
\vspace{1mm}

\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob knows that Amy knows whether \\ \hspace{3mm} it's tails.  \\
\end{tabular}
\end{example}

\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

@@

% $\tuple{K,+,\cdot,*,\bar{},0,1,{}^c,\Diamond_1,...,\Diamond_n}$

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of worlds
from the example.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St | Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}
\end{tcolorbox}
\caption{\raggedright Definition in Fst of KAT product and KAT Kleene star.
  Where {\tt X} and {\tt Y} are regular sets and
  {\tt R} and {\tt S} are regular relatiions,
  {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  {\tt X|Y} is the union of {\tt X} and {\tt Y},
  {\tt \~{}X} is the complement of {\tt X},
  {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  {\tt R{.}l} is the co-domain of {\tt R}, and
  {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

\begin{example}   \label{gs1fst}
  \renewcommand{\arraystretch}{1.2}
Worlds coded as strings
\vspace{1mm}

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 1 0} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

English sentences are mapped to terms in the logical language via an interpreted
grammar, and these terms are in turn interpreted as propositions (sets of
possible worlds).  The grammar is a semantically interpreted multi-modal
categorial grammar, consisting of a lexicon of words, their categorial types,
and interpretations in a logical lambda language.  The grammar covers basic
statives (it's heads), that- and whether-complements of {\em know}, sentence
negation, and predicate and sentence conjunction.  Figure \ref{lexiconfigure}
gives illustrative lexical entries.\footnote{Category symbols use
  Lambek/Bar-Hillel notation for slashes, so that $(d \backslash t)\slash(d
  \backslash_D t)$ combines with $d \backslash_D t$ on the right to give a value
  that combines with $d$ on the left to give $t$.}\footnote{Lambda abstractions
  with multiple parameters are written $\lambda x~y.~e$ rather than the more verbose
  $\lambda x.\lambda y.e$.} The grammar and semantics are optimized for a simple
fragment of English concerned with clausal complementation.  The agent names
{\em Amy} and {\em Bob} contribute the epistemic alternative relations for those
agents, rather than individuals.  This is possible because the agents are never
arguments of extensional predicates.  The root verb {\em know} contributes
existential modal force.  The complementizers {\em that} and {\em whether} are
the heads of their dominating clauses, and assemble an alternative relation,
modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to express
universal modal force.

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & ${R}_a$\\
        Bob & $e$ & ${R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
      \end{tabular}
  \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a logical language that extends the Epik term language with lambda. }
  \label{lexiconfigure}
\end{figure}

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are
used to control the derivation.
%
%For instance the category of {\em
%  heads} is $d \backslash_D t$.  The dummy expletive subject {\em it}
%has category $d$.  However the phrase {\em it heads} of category $t$
%can not be formed, because $\backslash_D$ is not syntactically active
%as a function.  Instead {\em it is heads} is formed with a predicator
%{\em is} of category $(d \backslash t)\slash(d \backslash_D t)$.
%Similarly {\em knows} has a category with the top-level slash
%$\slash\!_M$, and combines to form a sentence as an argument of {\em
%  that} or {\em whether}. These complementizers have a category that
%looks for the category of {\em know} on the left, after combining with
%a complement sentence on the right.
The semantic translations in the third column of Figure \ref{lexiconfigure} use
the Epik term language, incremented with lambda.  The body of $\lambda
x.\bullet^*h$, which is the semantic lexical entry for {\em heads}, is a
term denoting the set of all worlds where the coin is heads, expressed as the
set of all guarded strings that end with a Boolean valuation where the primitive
proposition $h$ (it's heads) is true.
%There is $\lambda x$ at the
%front because the grammar formalism enforces a strict correspondence
%between syntactic and semantic types. However, the lambda does not
%bind anything, because sentences such as {\em it isn't heads} have an
%expletive subject.
The body of $\lambda p.\lambda R. {\Diamond}_R p$, which is the semantic lexical
entry of {\em knows}, is an Epik term denoting the pre-image of the
world-alternative relation contributed by the subject.  This is not the right
semantics for {\em Amy knows that it's heads}, because it is an existential
modality ${\Diamond_R}p$, rather than an universal modality ${\Box_R}p$.  This
is corrected by the complementizer {\em that} or {\em whether}, which introduces
the dual.

Sentences are parsed with a chart parser for categorial grammar.  The semantics
for complex phrases are obtained by syntactic application of semantic
translations, accompanied by beta reduction.  In consequence, the semantic term
translating a sentence is a term of the logical language language.  Such a term
designates a set of possible words (guarded strings) in the possible worlds
model deterimened by an Epik specification.  By way of example, \exi{three}{a}
is an English sentence with predicate conjunction and three levels of clausal
embedding.  Using the grammar and parser, the sentence is mapped to the Epik
term in \exi{three}{b}.

\begin{examples} \label{three}
\itema Amy knows that Bob knows that Amy knows whether it is heads and knows that Bob doesnt know that Amy knows that it is tails.
%% \itemb {\tt (And (Not (Diamond amy (Not (Not (Diamond bob (Not (Or (Not (Diamond amy (Not heads))) (Not (Diamond amy heads))))))))) (Not (Diamond amy (Not (Not (Not (Diamond bob (Not (Not (Diamond amy (Not tails))))))))))) }
%% \itemc
%% \(\neg \Diamond_{\text{Amy}} \neg \neg \Diamond_{\text{Bob}} \neg \neg (\Diamond_{\text{Amy}}\bar h + \neg \Diamond_{\text{amy}} h) \cdot
%% \neg \Diamond_{\text{Amy}} \neg \neg \neg \Diamond_{\text{Bob}} \neg \neg \Diamond_{\text{Amy}} \bar t\)
\itemb \(\displaystyle \Box_{\text{Amy}}\Box_{\text{Bob}}(\Box_{\text{Amy}} \neg \bullet^* h + \Box_{\text{Amy}}\neg \bullet^* t)\) \hfill \mbox{} \(~~\cdot~~\Box_{\text{Amy}}\neg\Box_{\text{Bob}}\Box_{\text{Amy}}\neg \bullet^* t\)

\end{examples}

\exi{three}{b} is compiled an Fst finite state machine with -- nodes and
-- edges that accepts a countably infinite set of strings (worlds). Or similarly
interpreted as a lazily computation of guarded strings.  These are concrete
computational representations of the proposition denoted by \exi{three}{a}.
This contrasts with standard computational approaches to intensional semantics,
sentences are translated into logical formulas that have a mathematical
interpretation as a set of worlds, but not a computational one.  Our
representations can be used to check entailment between sentences (via a subset
check), print random worlds that satisfy the sentence (Fst only), print all
worlds shorter than some bound, and the like.

\section{Conclusion}

\ehc{Insert happy words!}

\pagebreak

>>>>

\newcommand{\ehc}[1]{\textit{\textcolor{red}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}

\newcommand{\K}{\mathcal{K}}
\newcommand{\B}{\mathcal{B}}

@@

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o1  h:h \\
            event o0  t:t \\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent amy \\
            \hspace{3mm} o1  -> o1 \\
            \hspace{3mm} o0  -> o0 \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

@@

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{llll}
      \multicolumn{4}{l}{\text{events}~~e\in \mathsf{P}} \\
      p,q &\multicolumn{3}{l}{::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap} \\
      \Box_a p & \triangleq \neg\Diamond_a\neg p &
      \bullet & \triangleq \sum_{e \in \mathsf P}e \\      
      p \wedge q &  \triangleq \neg(\neg p + \neg q) &
      p \to q & \neg p + q      
      \end{array} \)
  \end{minipage}}

@@

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
\ehc{I took the `that's out of these sentences. I remember from class
  that even if they are assigned null in the PF they still exist in
  the LF. Is it okay to leave it out here?}

\subsection{Standard Epistemic Modalities}

Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$ into
Epik\footnote{The choice to refer to $\Diamond_a$ as an ``epistemic''
  primitive rather than a ``doxastic'' primitive, is that it is
  essential in constructing both the doxastic and epistemic
  operators. A ``doxastic'' moniker would suggest the
  unrepresentability of epistemic modalities}. These encodings are
shown below:

\begin{example}\label{encodings}
  \( \begin{array}[t]{ccl}
    \textsc{Belief} & \B_a~p &\triangleq \Box_a~p \\
    \textsc{Knowledge} & \K_a~p & \triangleq p \wedge \B_a~p
    \end{array}\)
\end{example}

%% As a consequence of these encodings, the standard dualities
%% hold. These are shown in \ex{modalprops}.
%% \begin{example} \label{modalprops}
%%   \(\K_a~p \to \B_a~p \qquad \B_a~p \to \neg\K_a\neg p \)
%% \end{example}

To construct different types of reasoners (e.g. accurate, inaccurate,
etc), users of Epik can construct an agents alternative relation
accordingly. Observe the discussion of modal axioms \textbf{T} and
\textbf{D} in Section~\ref{sec:axioms}.

\subsection{Guarded String Interpretation}
\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

@@

\subsection{Axiomatic Classification}
\label{sec:axioms}

To situate our logic as a modal logic, we investigate the soundness of
the standard modal axioms~\cite{????} in Epik's semantics. Some of
these standard axioms (\ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

\begin{example}\label{valid-axioms}
  \begin{enumerate}[align=left]
  \item[\textbf{N}] If $p$ is valid, then $\Box_ap$ is valid
  \item[\textbf{K}] $\Box_a(p \to q) \to \Box_a~p \to \Box_a~q$ is valid.
  \end{enumerate}
\end{example}

\begin{example}\label{conditionally-valid-axioms}
  \(\begin{array}[t]{lll}
  \textbf{T} & \Box_a p \to p & \text{if } g\mathrel{\hat{R}_a} g, \forall g \\
  \textbf{D} & \Box_a p \to \Diamond_a p & \text{if } g \in \dom(\hat{R}_a), \forall g \\
  \textbf{4} & \Box_a p \to \Box_a\Box_a p & \text{if } \hat{R}_a \text{ idempotent}
  \end{array}\)
\end{example}

The condition on \textbf{4} is essentially just a restatement of the
theorem in relational terms. The conditions on the remaining standard
modal axioms, \textbf{B} and \textbf{5}, are of a similarly trivial
flavor.

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of worlds $w_2$ and $w_3$
from the example.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St | Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}
\end{tcolorbox}
\caption{\raggedright Definition in Fst of KAT product and star.
  %% Where {\tt X} and {\tt Y} are regular sets and
  %% {\tt R} and {\tt S} are regular relatiions,
  %% {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  %% {\tt X|Y} is the union of {\tt X} and {\tt Y},
  %% {\tt \~{}X} is the complement of {\tt X},
  %% {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  %% {\tt R{.}l} is the co-domain of {\tt R}, and
  %% {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

\begin{example}   \label{gs1fst}
  \renewcommand{\arraystretch}{1.2}
\begin{tabular}[t]{cl}
%% {\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
%% $w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 1 0} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

English sentences are mapped to terms in the logical language via a
semantically interpreted multi-modal categorial grammar, consisting of
a lexicon of words, their categorial types, and interpretations in a
logical lambda language.  The grammar covers basic statives ({\em it's
  heads}), {\em that-} and {\em whether}-complements of {\em know}, predicate and
sentence negation, and predicate and sentence conjunction.  Figure
\ref{lexiconfigure} gives illustrative lexical
entries.\footnote{Category symbols use Lambek/Bar-Hillel notation for
  slashes, so that $(d \backslash t)\slash(d \backslash_D t)$ combines
  with $d \backslash_D t$ on the right to give a value that combines
  with $d$ on the left to give $t$.}\footnote{Lambda abstractions with
  multiple parameters are written $\lambda x~y.~e$ rather than the
  more verbose $\lambda x.\lambda y.e$.} The grammar and semantics are
optimized for a simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute
the epistemic alternative relations for those agents, rather than
individuals.  The root verb {\em know} contributes existential modal
force.  The complementizers {\em that} and {\em whether} are the heads
of their dominating clauses, and assemble an alternative relation,
modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to
express universal modal force.

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & ${R}_a$\\
        Bob & $e$ & ${R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
      \end{tabular}
  \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a logical language that extends the Epik term language with lambda. }
  \label{lexiconfigure}
\end{figure}

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are
used to control the derivation.
%
%For instance the category of {\em
%  heads} is $d \backslash_D t$.  The dummy expletive subject {\em it}
%has category $d$.  However the phrase {\em it heads} of category $t$
%can not be formed, because $\backslash_D$ is not syntactically active
%as a function.  Instead {\em it is heads} is formed with a predicator
%{\em is} of category $(d \backslash t)\slash(d \backslash_D t)$.
%Similarly {\em knows} has a category with the top-level slash
%$\slash\!_M$, and combines to form a sentence as an argument of {\em
%  that} or {\em whether}. These complementizers have a category that
%looks for the category of {\em know} on the left, after combining with
%a complement sentence on the right.
The semantic translations in the third column of Figure \ref{lexiconfigure} use
the logical language, incremented with lambda.  The body of $\lambda
x.\bullet^*h$, which is the semantic lexical entry for {\em heads}, is a
term denoting the set of all worlds where the coin is heads, expressed as the
set of all guarded strings that end with a Boolean valuation where the primitive
proposition $h$ (it's heads) is true.
%There is $\lambda x$ at the
%front because the grammar formalism enforces a strict correspondence
%between syntactic and semantic types. However, the lambda does not
%bind anything, because sentences such as {\em it isn't heads} have an
%expletive subject.
The body of $\lambda p.\lambda R. {\Diamond}_R p$, which is the semantic lexical
entry of {\em knows}, is an term denoting the pre-image of the
world-alternative relation contributed by the subject.  This is not the right
semantics for {\em Amy knows that it's heads}, because it is an existential
modality ${\Diamond_R}p$, rather than an universal modality ${\Box_R}p$.  This
is corrected by the complementizer {\em that} or {\em whether}, which introduces
the dual.

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by application of semantic
translations, accompanied by beta reductions that eliminate all
lambdas in logical forms for clauses.  In consequence, the semantic
term translating a sentence is a term of the logical language.  Such a
term designates a set of possible words (guarded strings) in the
possible worlds model determined by an Epik specification.  By way of
example, \exi{three}{a} is an English sentence with conjunction and
several levels of clausal embedding.  Using the grammar and parser,
the sentence is mapped to the term in \exi{three}{b}. Then
\exi{three}{c} shows a simplified logical from constructed form
\exi{three}{b} using simply logical equivalences. The negations are
propositional. The double negation, while logically redundant, comes
syntactically from the grammar. This term is compiled in an
implementation of the finite state calculus to a finite state machine
with six states and eight arcs, which accepts a countably infinite set
of worlds.  In this way the methodology ``directly'' represents the
set of worlds denoted by \exi{three}{a}.

\newcommand{\HEADS}{\bullet^*h}
\newcommand{\AKWH}{\neg (\neg \HEADS \vee \Diamond_{a} \neg\HEADS) + \neg(\HEADS \wedge \Diamond_{a}\HEADS)}
\newcommand{\BKAKWH}{\neg(\neg(\AKWH) + \Diamond_{b}\neg(\AKWH))}
\newcommand{\AKBKAKWH}{\neg(\neg(\BKAKWH) + \Diamond_{a}\neg(\BKAKWH))}

\begin{examples} \label{three}
\itema It's tails and Amy knows that Bob knows that Amy knows whether it's heads.
% (_*;t) & ~ <amy> ~ ~ <bob> ~((~ (<amy> ~ (_*;h))) + ~ (<amy> (_*;h)) )
\itemb \( \displaystyle \bullet^*t \wedge \AKBKAKWH \)\\
\itemc \(\bullet^*t \wedge \K_{a}(\K_{b}(\K_{a} \bullet^* h + \K_{a} \neg \bullet ^* h))\)
\itemd Amy knows that it's tails.
\iteme \(
\bullet ^*t \wedge \neg \Diamond_{a} \neg \bullet^*t~~~~~~(\equiv \K_a \bullet^* t)
\)

\end{examples}
Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by
the grammar.  Logical relations between propositions are checked in
the finite state calculus by checking set-theoretic relations between
sets of worlds.  In this case, the propositions \exi{three}{b} and
\exi{three}{e} are equivalent, in the sense that they denote the same
set of worlds.  This equivalence is attested computationally in the
possible worlds model.

%% \begin{examples} \label{entail}
%% \itema Amy knows that it's tails.
%% \itemb \(
%% \neg \Diamond_{a} \neg \bullet^*t
%% \)
%% % \itemc \(\displaystyle \Box_{a}\Box_{b}(\Box_{a} \neg \bullet^* h + \Box_{a}\neg \bullet^* t)\) \hfill \mbox{} \(~~\cdot~~\Box_{a}\neg\Box_{b}\Box_{a}\neg \bullet^* t\)
%% \end{examples}

\section{Discussion}
The methodology presented here is designed for use in research in
linguistic semantics, and for education at the level of a second
graduate course in formal semantics, covering intensionality.  There
are straighforward extensions to additional semantic phenomena, such
as tense and perfective aspect as in \exi{extension}{a}, and the combination of
metaphysical modality and prospective aspect in \exi{extension}{b}.

\begin{examples}   \label{extension}
  \itema Amy has learned that Bob had learned that it's heads.
  \itemb Amy might learn that it's heads.
\end{examples}

The development here was concerned with defining concrete possible
worlds models, and applying them in natural language semantics.  An
issue for further investigation are the mathematical characterizations
of epistemic KATs, e.g. sound and complete axioms, coalgebra, decidability.
